---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
# Download the file if necessary and load it
if (!file.exists("activity.csv")) {
    fileUrl <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(fileUrl, destfile = "activity.zip", method = "curl")
    unzip("activity.zip")
}
activity <- read.csv("activity.csv")
str(activity)
```

## What is mean total number of steps taken per day?
```{r}
activity$date <- as.Date(activity$date)
steps_sum_per_day <- tapply(activity$steps,activity$date,sum, na.rm=T)
steps_sum_per_day

hist(steps_sum_per_day,breaks=20,main="Total Number of Steps per Day",xlab="Number of Steps per Day")

# Calculate and display mean and median number of steps per day
mean_total_steps_per_day <- mean(steps_sum_per_day)
mean_total_steps_per_day
median_total_steps_per_day <- median(steps_sum_per_day)
median_total_steps_per_day
```

## What is the average daily activity pattern?
```{r}
require(ggplot2)
steps_mean_per_interval <- as.numeric(tapply(activity$steps,activity$interval,mean, na.rm=T))

interval <- unique(activity$interval)
interval_mean_df<-data.frame(interval,steps_mean_per_interval)
ggplot(interval_mean_df,aes(x=interval,y=steps_mean_per_interval)) +     geom_line()+scale_x_continuous(name="interval",breaks=c(0,500,800,1000,1200,1600,2000,2400)) + ylab("Mean of Daily Steps per Interval")
max_mean <- max(steps_mean_per_interval)
max_interval <- interval[which(steps_mean_per_interval==max(steps_mean_per_interval))]
```
NOTE: The maximum number of steps **`r max_mean`** occurs at **`r max_interval`** interval.

## Imputing missing values

#### Strategy for Imputing missing values ####
The values for the number of steps per day is not completely random since there is a pattern to human activity. I used the average per interval to impute the NA values to enhance rather than attenuate the existing pattern if there is any.  
```{r}
# Number of missing values
na_count <- sum(is.na(activity$steps))
na_count
# Create a new data set with missing values filled in with the mean for the 5-minute interval
new_activity <- activity
# create a vector of indices with missing steps values
na_idx <- which(is.na(new_activity$steps))
# create a vector of matches interval_mean_df as the lookup table
na_matches <- match(new_activity[na_idx,"interval"],interval_mean_df$interval)
# fill in the missing values with the mean
new_activity$steps[na_idx] <- interval_mean_df[na_matches,"steps_mean_per_interval"]
# Number of missing values in new_activity
new_na_count <- sum(is.na(new_activity$steps))
new_na_count
# calculate sum, mean and median for the new_activity
new_steps_sum_per_day <- tapply(new_activity$steps,new_activity$date,sum, na.rm=T)
new_mean_total_steps_per_day <- mean(new_steps_sum_per_day)
new_median_total_steps_per_day <- median(new_steps_sum_per_day)

# Plot the histograms comparing the two
par(mfrow=c(1,2))
hist(steps_sum_per_day,breaks=20,ylim=c(0,20),xlab="Sum of steps per day",main="Histogram with NAs")
hist(new_steps_sum_per_day,breaks=20,ylim=c(0,20),xlab="Sum of steps per day",main="Histogram with Imputed NAs")

# Display mean and median number of steps per day with imputed NAs
new_mean_total_steps_per_day
new_median_total_steps_per_day
```

**Filling the NAs with the mean for each interval aligned the mean and the median of the total steps per day.**

## Are there differences in activity patterns between weekdays and weekends?
```{r}
require(lattice)
type_of_day<-ifelse(weekdays(new_activity$date) %in% c("Saturday","Sunday"),"weekend","weekday")

new_activity <- cbind(new_activity,type_of_day)
weekday_subset <- subset(new_activity,type_of_day=="weekday")
weekday_mean <- as.numeric(tapply(weekday_subset$steps,weekday_subset$interval,mean, na.rm=T))
weekend_subset <- subset(new_activity,type_of_day=="weekend")
weekend_mean <- as.numeric(tapply(weekend_subset$steps,weekend_subset$interval,mean, na.rm=T))

weekday_df <- data.frame(mean=weekday_mean,day=c("weekday"),interval)
weekend_df <- data.frame(mean=weekend_mean,day=c("weekend"),interval)
all_days <- rbind(weekday_df,weekend_df)

xyplot(mean ~ interval | day,data=all_days,type="l",layout=c(1,2),xlab="Interval",ylab="Number of steps")

# weekday versus weekend
weekday_max_mean <- max(weekday_df$mean)
weekday_max_interval <- weekday_df$interval[which(weekday_df$mean==weekday_max_mean)]
weekend_max_mean <- max(weekend_df$mean)
weekend_max_interval <- weekend_df$interval[which(weekend_df$mean==weekend_max_mean)]
```
The weekday maximum number of steps **`r weekday_max_mean`** occurs at **`r weekday_max_interval`** interval while weekend is **`r weekend_max_mean`** at **`r weekend_max_interval`** interval.
